Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-10-22 paper/main.tex
start working on pldi intro

In software engineering and software verification, we often would like
to have static information about the quantitative behavior of
programs.  For example, stack and heap-space bounds are important to
ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
% Clock-cycle bounds are needed to guarantee the safety of real-time
% systems [XXX wilhelm].  
Static energy usage information is critical for autonomous systems and
has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
Worst-case time bounds can help to create constant-time
implementations that prevent side-channel
attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
bounds are used to ensure the accuracy of programs that are executed
on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
needed to verify cryptographic protocols~\cite{BartheGB09}.
% XXX Mention differential privacy here?
%
In general, quantitative performance information at design time can
provide useful feedback for developers.

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of existing techniques in practice:
\itemskip
\begin{itemize}
\item Analysis tools are black boxes that\iffull{either}{} deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\itemskipIn
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\itemskipIn
\item Existing techniques often rely on complex external tools such as
  abstract interpretation--based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
\itemskip
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\itemskip
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-4ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
\itemskip
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation.  The choice of C is
primarily motivated by \iffull{our}{the} ongoing work on the formally verified
hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
developed in C and is supposed to provide verified guaranties on
timing and memory usage.  Moreover, C is a natural choice because it
is still the most widely used language for system development; in
particular in safety-critical embedded and real-time systems where
resource bound analyses are often required by regulatory
authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
%
\itemskip
\begin{enumerate}
\item We define an operational cost semantics for CompCert Clight that
  defines the resource consumption of terminating and diverging
  executions.  The cost is parametric in a user-definable cost metric
  and can be negative if resources are released during an execution.

\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\itemskipIn
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver, and does not require any
  fixpoint computations to obtain loop invariants.
\itemskipIn
\item We show with a publicly available prototype implementation, and
  an experimental evaluation, that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
\itemskip
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still supporting
automation that sometimes goes beyond the capabilities of existing
techniques.  Furthermore, the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during execution.

Our starting point is a recent work~\cite{veristack14} \iffull{in
  which we have}{that has} formally verified bounds on the stack usage
of C programs.  A key part of this stack-bound verification is based
on a quantitative Hoare logic for deriving abstract \emph{stack
  bounds} for CompCert Clight programs\ifshort{.}{ that depend on the
  depth of (recursive) function calls.}  % For
% non-recursive programs it is possible to compute derivations for such
% stack bounds in the quantitative Hoare logic automatically.  
% By compiling these bounds together with the C program, we have derived
% verified stack bounds for x86 assembly code.
%
Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  % The
% quantitative logic is implemented in Coq and proved sound with respect
% to a small-step operational cost semantics for Clight.
% that counts a
% constant resource consumption for each atomic execution step.  The
% resource consumption at each step is parametric in a user-definable cost
% metric and can be negative to express the release of resources (as
% needed, e.g., to reason about memory usage).  Furthermore, a specific
% \emph{tick} function can be inserted in the source code to further
% customize the cost model.
In the implementation of the logic, we use
a shallow embedding in Coq which makes the logic very flexible.  There
is practically no limitation on the format of the derived resource
bounds since they can be any function that is definable in Coq.  The
derived bounds can also be parametric in a set of cost metrics or
specifically apply to one fixed cost metric.

The precision and expressivity of the quantitative Hoare logic provide
an ample foundation for reasoning about resource consumption of Clight
programs.  % The logic is naturally compositional and the derived
% \emph{quantitative Hoare triples} can be used to specify the resource
% consumption of library functions or to describe resource contracts
% between different functions. 
However, reasoning about quantitative
properties can often be more tedious than reasoning about intentional
properties.  Consequently, automation is inevitable to derive resource
bounds for large code bases like CertiKOS.  Such an automation was
easily achievable in the case of constant stack bounds for a code base
that does not contain recursive functions~\cite{veristack14}.  In
general however, resource bounds have to be parametric in the
arguments of a function and depend on the number of loop iterations
and (recursive) function calls that are performed by the function.

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been an inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  Type-based amortized resource analysis works
well for functional programs with pattern matching but it has been a
long-time open problem to apply it to C-like programs with control
flow that depends on integer arithmetic, negative numbers, and non-linear
control flow.
% example, it is a
% long-time open problem how to use type-based amortized analysis to
% derive bounds for functions whose resource consumption depends on
% (possibly negative) integers and non-sequential control flow as
% introduced by \code{break} and \code{return} statements.
% Unfortunately, these are the functions \iffull{with non-constant resource
% consumption}{} that are most common in CertiKOS and other system
% software.

Our \emph{main contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers. It also handles programs with mutually recursive functions
as well as \code{break} and \code{return} statements.  The
potential-based approach of amortized analysis provides a single
mechanism for analyzing programs that need special treatment in other
techniques; including
\itemskip
\begin{itemize}
\item interaction between \emph{sequential loops or function calls} through
  size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each others number of iterations,
\itemskipIn
\item and \emph{tricky iteration patterns} as found for instance in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
%
The main innovation that makes the analysis practical for C is the use
of interval sizes in potential functions instead of the sizes of
variables.  This leads to precise bounds of programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract interpretation--based methods such as disjunctive invariant
generation~\cite{GulwaniMC09} and symbolic backward
execution~\cite{GulwaniZ10}.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects the resource cost and size changes of
integer variables in the program; a solution of the linear program
immediately yields a resource-usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

Following the development steps of automatic amortized analysis for
functional programs~\cite{Jost03,HoffmannH10}, we deliberately
restrict ourselves to linear bounds in the automatic analysis (there
are no restrictions for the manually-derived bounds in the logic).
More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
where $a$ and $b$ are integer variables or constants.  The reason for
our focus on linear bounds is mainly clarity of presentation.  We
already experimented with an extension to multivariate resource
polynomials~\cite{HoffmannAH11,HoffmannS13} but this would make the inference
rules considerably more involved and should better be described
separately.  However, we developed the linear inference system so that
the extension to polynomial bounds shall work smoothly.

We have evaluated the automatic analysis with system code and examples
from the literature.  \iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicality and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds a maximal element in an array, and a linear
bound that amortizes the cost of $k$ successive increments to a binary
counter.

% To show the practicality of our approach we developed a case study
% with different real-time schedulers implemented in C.  To perform a
% schedulability analysis that takes into account the overhead of the
% scheduler we derived time bounds for context switching and
% initialization of the scheduler.  We then computed time bounds for a
% set of given user task and took account the previously computed
% scheduler overhead to determine whether the scheduling requirements of
% all user tasks can be met.

% With the intention, of making the material easily accessible we first
% informally describe the new automatic amortized analysis
% (\pref{sec:inform}).  We then formalize the intuition with the
% operational cost semantics an the inference rules for the amortized
% analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
% quantitative Hoare logic and its soundness proof that is formalized in
% Coq.  In \pref{sec:inter} we prove the soundness of the automatic
% amortized analysis by showing that derivation using the inference
% rules can be seen as derivations in the quantitative logic.  We also
% show how to integrate hand-derived bounds with automatically-derived
% bounds.  \pref{sec:exper} contains the results of the experimental
% evaluation and \pref{sec:related} describes related research.

>>>>

In software engineering and software verification,
we often would like to have static information
about the quantitative behavior of programs.
For example, stack and heap-space bounds
are important to ensure the reliability of
safety-critical systems~\cite{veristack14,Regehr05}.
Static energy usage information is critical
for autonomous systems and has applications in
cloud computing~\cite{CohenZSL12,CarrollH10}.
Worst-case time bounds can help to create
constant-time implementations that prevent
side-channel attacks~\cite{KasperS09,BartheBCLP14}.
Loop and recursion-depth bounds are used to
ensure the accuracy of programs that are executed
on unreliable hardware~\cite{CarbinMR13} and
complexity bounds are needed to verify cryptographic
protocols~\cite{BartheGB09}.  In general, quantitative
performance information at design time can provide useful
feedback for developers.

Static analysis of quantitative properties of imperative
programs is an active research area and recent years
have seen many innovations.  Notable tools that
have been developed include SPEED~\cite{GulwaniMC09},
KoAT~\cite{BrockschmidtEFFG14}, PUBS~\cite{AlbertAGPZ12},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.

These tools can derive impressive results on realistic
system software.  However, there are still theoritical
shortcomings that remain open problems to this day:

\begin{itemize}
\item The computed bounds are often
  \emph{non-compositional} local bounds (for a single
  (nested) loop) that are difficult to combine to
  global whole program bounds.
\itemskipIn
\item When an automatic tool fails on a loop of the
  program, no user interaction is possible to derive
  a bound on the full program.
\itemskipIn
\item Existing tools work on a highly abstracted
  version of the program, thus, are unable to provide
  a simple syntactic proof of the correctness of their
  bound.
\itemskipIn
\item Maybe we can mention resource freeing                    XXX
\end{itemize}

This paper presents a new framework to derive worst-case
resource bounds for C programs.  Our choice of C is
mainly motivated by its strong presence in system and
embedded software.  These applications often have to meet
strict time and memory constraints.  To ensure that these
requirements are met, profiling is not enough because non
exhaustive.  In contrast, static analysis tools provide
stronger and sound guarantees.

Our compositional framework is built of three parts: 1-
a clear cost model for C programs based on instrumented
semantics; 2- a quantitative Hoare logic allowing
compositional reasoning on the resource consumption of
C programs at a syntactic level; 3- an automated tool
that derives bounds in the previous logic.

Compared to existing work, our framework is a jump in
compositionality and correctness: the analysis is no
longer centered on single loops but on complete programs,
a mix of automation and interaction can be used to derive
global bounds, and bounds are computed and proved within
a clear syntactic deductive system.  Compositionality is
due first to the logic that soundly allows mixes of hand
derived bounds and automatically generated bounds.  It is
also due to the automated tool that keeps track of loop
bounds and size changes of variables in a unified system.

The cost model and the quantitative Hoare logic
presented are based on a recent work~\cite{veristack14}
that formally verifies bounds on the stack usage of
C programs.  We generalize this work by allowing to
reason on any resource.  The automation, which is the
keystone of this paper, is based on amortized analysis
techniques~\cite{jan-something}.  It makes the analysis
of large programs tractable by automatically generating
resource bounds in the previous logic on a large class
of programs.

Our automated analysis is able to infer linear
resource bounds on C programs with integer loops.
To our knowledge this is the first tool based on
amortized analysis able to deal with negative numbers and
differences of variables.  It is also the first resource
analysis tool for C that deals naturally with recursive
functions, resources that can be freed (e.g. memory),
and sequenced loops.  Automated tools for functional
programs already existed~\cite{Jost03,HoffmannAH12}
but it has been a long-time open problem to apply it to
C-like programming languages with control flow depending
on integer arithmetic.

Compared to more classical approaches based on ranking
functions, tools that are based on amortized analysis
only need one mechanism to provide features that are
usually treated non uniformly:

\begin{itemize}
\item interaction between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\item and \emph{amortized bounds} as found
  in the Knuth-Morris-Pratt algorithm for string
  search.
\end{itemize}

This paper presents a compositional framework for
deriving certified resource bounds on C programs. We
emphasize the following contributions: 1- an automatic
analysis that generates linear compositional and
certified resource bounds; 2- an infrastructure verified
in Coq that lets a programmer derive complex free-form
resource bounds on complete C programs.

Give a description of the sections here.                    XXX

% In software engineering and software verification, we often would like
% to have static information about the quantitative behavior of
% programs.  For example, stack and heap-space bounds are important to
% ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
% % Clock-cycle bounds are needed to guarantee the safety of real-time
% % systems [XXX wilhelm].
% Static energy usage information is critical for autonomous systems and
% has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
% Worst-case time bounds can help to create constant-time
% implementations that prevent side-channel
% attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
% bounds are used to ensure the accuracy of programs that are executed
% on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
% needed to verify cryptographic protocols~\cite{BartheGB09}.
% % XXX Mention differential privacy here?
% %
% In general, quantitative performance information at design time can
% provide useful feedback for developers.
% Static analysis of quantitative properties of imperative programs is
% an active research area and recent years have seen many innovations.
% Notable tools that have been developed include
% SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
% PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
% %
% While the these tools can derive impressive results for realistic
% software, there still exist shortcomings that hamper the application
% of existing techniques in practice:
% \itemskip
% \begin{itemize}
% \item Analysis tools are black boxes that\iffull{either}{} deliver a result or fail
%   without enabling \emph{user interaction} to manually or
%   semi-automatically derive bounds for challenging parts of the
%   program.
% \itemskipIn
% \item The computed bounds are often \emph{non-compositional} local
%   bounds (for a single (nested) loop) that are difficult to combine to
%   global whole program bounds.
% \itemskipIn
% \item Existing techniques often rely on complex external tools such as
%   abstract interpretation--based invariant
%   generation~\cite{GulwaniMC09} or translation of the program into a
%   term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
%   providing \emph{verifiable certificates} for the correctness of the
%   derived bound.
% \end{itemize}
% \itemskip
% %
% While there has been much progress in static quantitative analysis,
% Knuth correctly points out in a recent interview~\cite{KnuthInter}
% that the state-of-the-art in formal quantitative methods still falls
% short in comparison with semantic techniques.
% %
% \itemskip
% \begin{quote}
%   % Consider, for example, a computer program that sorts a list of
%   % numbers into order.
%   $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
%   we have formal definitions of semantics, and tools by which we can
%   verify that sorting is indeed always achieved. My job is to go
%   beyond correctness, to an analysis of such things as the program's
%   running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
%   describes the program's performance, and all of my colleagues agree
%   with me that the recurrence is "obviously" valid. Yet I have no
%   formal tools by which I can prove that my recurrence is right. I
%   don't really understand my reasoning processes at all!
%   \vspace{-4ex}
%   \begin{flushright}
%     -- Donald E. Knuth, 2014
%   \end{flushright}
% \end{quote}
% \itemskip
% %
% In this work, we develop a resource analysis framework for C programs
% that is based on a solid semantic foundation.  The choice of C is
% primarily motivated by \iffull{our}{the} ongoing work on the formally verified
% hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
% developed in C and is supposed to provide verified guaranties on
% timing and memory usage.  Moreover, C is a natural choice because it
% is still the most widely used language for system development; in
% particular in safety-critical embedded and real-time systems where
% resource bound analyses are often required by regulatory
% authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
% %
% \itemskip
% \begin{enumerate}
% \item We define an operational cost semantics for CompCert Clight that
%   defines the resource consumption of terminating and diverging
%   executions.  The cost is parametric in a user-definable cost metric
%   and can be negative if resources are released during an execution.
% \item We develop a quantitative Hoare logic for interactively deriving
%   resource bounds for Clight programs.  The logic is implemented and
%   proved sound in the Coq Proof Assistant with respect to the cost
%   semantics of Clight.
% \itemskipIn
% \item We describe an automatic static analysis that computes bounds
%   together with derivations in the quantitative Hoare logic.  The
%   analysis is inspired by type-based amortized resource analysis for
%   functional programs~\cite{Jost03,HoffmannAH11}.  It computes
%   derivations by generating simple linear constraints that can be
%   solved by an off-the-shelf LP solver, and does not require any
%   fixpoint computations to obtain loop invariants.
% \itemskipIn
% \item We show with a publicly available prototype implementation, and
%   an experimental evaluation, that our novel amortized resource
%   analysis works precisely and efficiently for challenging loop and
%   recursion patterns, and that the derived constant factors in the
%   bounds are close to or identical with the optimal ones.
% % \item We prove the completeness of the program logic
% %   with respect to the operational cost semantics.
% \end{enumerate}
% \itemskip
% %
% To the best of our knowledge, this article presents the first resource
% analysis framework for C that makes it possible to combine non-trivial
% automatically derived bounds with interactively derived bounds in a
% proof system that produces verifiable certificates for the bounds.
% Our approach complements existing work since it provides a semantic
% foundation for the computation of bounds while still supporting
% automation that sometimes goes beyond the capabilities of existing
% techniques.  Furthermore, the automatic amortized analysis and the
% quantitative logic are naturally compositional and describe the
% resource behavior of code fragments without referring to the source
% code.  As a result, we directly derive global bounds that are
% functions of the input parameters of the program.  Another unique
% feature of our system is, that it can handle resources like memory
% that may become available during execution.
% Our starting point is a recent work~\cite{veristack14} \iffull{in
%   which we have}{that has} formally verified bounds on the stack usage
% of C programs.  A key part of this stack-bound verification is based
% on a quantitative Hoare logic for deriving abstract \emph{stack
%   bounds} for CompCert Clight programs\ifshort{.}{ that depend on the
%   depth of (recursive) function calls.}  % For
% % non-recursive programs it is possible to compute derivations for such
% % stack bounds in the quantitative Hoare logic automatically.
% % By compiling these bounds together with the C program, we have derived
% % verified stack bounds for x86 assembly code.
% %
% Our \emph{first contribution} is a generalization of the quantitative
% Hoare logic for CompCert Clight that enables us to derive resource
% bounds that are parametric in the resource of interest.  % The
% % quantitative logic is implemented in Coq and proved sound with respect
% % to a small-step operational cost semantics for Clight.
% % that counts a
% % constant resource consumption for each atomic execution step.  The
% % resource consumption at each step is parametric in a user-definable cost
% % metric and can be negative to express the release of resources (as
% % needed, e.g., to reason about memory usage).  Furthermore, a specific
% % \emph{tick} function can be inserted in the source code to further
% % customize the cost model.
% In the implementation of the logic, we use
% a shallow embedding in Coq which makes the logic very flexible.  There
% is practically no limitation on the format of the derived resource
% bounds since they can be any function that is definable in Coq.  The
% derived bounds can also be parametric in a set of cost metrics or
% specifically apply to one fixed cost metric.
% The precision and expressivity of the quantitative Hoare logic provide
% an ample foundation for reasoning about resource consumption of Clight
% programs.  % The logic is naturally compositional and the derived
% % \emph{quantitative Hoare triples} can be used to specify the resource
% % consumption of library functions or to describe resource contracts
% % between different functions.
% However, reasoning about quantitative
% properties can often be more tedious than reasoning about intentional
% properties.  Consequently, automation is inevitable to derive resource
% bounds for large code bases like CertiKOS.  Such an automation was
% easily achievable in the case of constant stack bounds for a code base
% that does not contain recursive functions~\cite{veristack14}.  In
% general however, resource bounds have to be parametric in the
% arguments of a function and depend on the number of loop iterations
% and (recursive) function calls that are performed by the function.
% For functional programs, there exist resource analysis systems---based
% on type-based amortized resource analysis~\cite{Jost03,
%   HoffmannAH12}---that can automatically derive complex polynomial
% bounds with tight constant factors.  This type-based amortized
% resource analysis has been an inspiration in the design of the
% quantitative Hoare logic and is therefore a natural candidate to
% automate the reasoning.  Type-based amortized resource analysis works
% well for functional programs with pattern matching but it has been a
% long-time open problem to apply it to C-like programs with control
% flow that depends on integer arithmetic, negative numbers, and non-linear
% control flow.
% % example, it is a
% % long-time open problem how to use type-based amortized analysis to
% % derive bounds for functions whose resource consumption depends on
% % (possibly negative) integers and non-sequential control flow as
% % introduced by \code{break} and \code{return} statements.
% % Unfortunately, these are the functions \iffull{with non-constant resource
% % consumption}{} that are most common in CertiKOS and other system
% % software.
% Our \emph{main contribution} is an automatic amortized resource
% analysis for CompCert Clight that computes derivations in the
% quantitative Hoare logic.  It is the first automatic amortized
% resource analysis that can derive bounds that depend on negative
% integers. It also handles programs with mutually recursive functions
% as well as \code{break} and \code{return} statements.  The
% potential-based approach of amortized analysis provides a single
% mechanism for analyzing programs that need special treatment in other
% techniques; including
% \itemskip
% \begin{itemize}
% \item interaction between \emph{sequential loops or function calls} through
%   size changes of variables,
% \itemskipIn
% \item \emph{nested loops} that influence each others number of iterations,
% \itemskipIn
% \item and \emph{tricky iteration patterns} as found for instance in
%   the Knuth-Morris-Pratt algorithm for string search.
% \end{itemize}
% \itemskip
% %
% The main innovation that makes the analysis practical for C is the use
% of interval sizes in potential functions instead of the sizes of
% variables.  This leads to precise bounds of programs whose resource
% consumption can be described as a function of sizes $|[x,y]| =
% \max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
% instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
%   i=i+K)} where the step-size $K>0$ is a constant.
% %
% Despite the apparent simplicity of the new analysis system, it is able
% to reproduce results from the literature (e.g,
% SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
% abstract interpretation--based methods such as disjunctive invariant
% generation~\cite{GulwaniMC09} and symbolic backward
% execution~\cite{GulwaniZ10}.  In contrast with abstract
% interpretation--based methods, our technique does not require any
% fixpoint computations to obtain loop invariants.  The mechanism we
% designed is able to leverage local assertions such as $x < y$ that we
% collect along the branching points of the program to obtain global
% resource invariants.  We achieve this by generating a linear
% constraint system that reflects the resource cost and size changes of
% integer variables in the program; a solution of the linear program
% immediately yields a resource-usage bound for the C
% program.  % We can naturally, handle
% % advanced control flow such as break, return, iteration over negative
% % integer variables, and mutually-recursive functions.
% Following the development steps of automatic amortized analysis for
% functional programs~\cite{Jost03,HoffmannH10}, we deliberately
% restrict ourselves to linear bounds in the automatic analysis (there
% are no restrictions for the manually-derived bounds in the logic).
% More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
% where $a$ and $b$ are integer variables or constants.  The reason for
% our focus on linear bounds is mainly clarity of presentation.  We
% already experimented with an extension to multivariate resource
% polynomials~\cite{HoffmannAH11,HoffmannS13} but this would make the inference
% rules considerably more involved and should better be described
% separately.  However, we developed the linear inference system so that
% the extension to polynomial bounds shall work smoothly.
% We have evaluated the automatic analysis with system code and examples
% from the literature.  \iffull{\pref{app:cat}}{The extended version
%   of this article~\cite{anon_extended}}{} contains more than $30$
% challenging loop and recursion patterns that we collected from open
% source software and the literature.  Our analysis can find
% asymptotically tight bounds for all but 1 of these patterns, and in
% most cases the derived constant factors are tight.  To compare our
% automatic analyzer with existing techniques, we tested our examples
% with tools such as KoAT~\cite{BrockschmidtEFFG14},
% Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
% show that the bounds that we derive are often more precise than those
% derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
% uses amortization techniques, is able to achieve a similar precision.
% %
% Several micro benchmarks demonstrate the practicality and
% expressiveness of the quantitative Hoare logic.  For example, we
% derive a logarithmic bound for a binary search function, a bound that
% depends on the contents of an array to describe the exact cost of a
% function that finds a maximal element in an array, and a linear
% bound that amortizes the cost of $k$ successive increments to a binary
% counter.
% % To show the practicality of our approach we developed a case study
% % with different real-time schedulers implemented in C.  To perform a
% % schedulability analysis that takes into account the overhead of the
% % scheduler we derived time bounds for context switching and
% % initialization of the scheduler.  We then computed time bounds for a
% % set of given user task and took account the previously computed
% % scheduler overhead to determine whether the scheduling requirements of
% % all user tasks can be met.
% % With the intention, of making the material easily accessible we first
% % informally describe the new automatic amortized analysis
% % (\pref{sec:inform}).  We then formalize the intuition with the
% % operational cost semantics an the inference rules for the amortized
% % analysis (\pref{sec:aa}).  In \pref{sec:logic} we describe the
% % quantitative Hoare logic and its soundness proof that is formalized in
% % Coq.  In \pref{sec:inter} we prove the soundness of the automatic
% % amortized analysis by showing that derivation using the inference
% % rules can be seen as derivations in the quantitative logic.  We also
% % show how to integrate hand-derived bounds with automatically-derived
% % bounds.  \pref{sec:exper} contains the results of the experimental
% % evaluation and \pref{sec:related} describes related research.
